---
title: "VAPB interactome"
author: "Birol Cabukusta"
date: "2026-07-28"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F)
```

# VAPB interactome
## 2026

Here we performed TurboID experiments with N-terminal TurboID-tagged VAPB, its point mutant unable to interact with FFAT motifs [K87D, M89D], TM helix only, soluble, free TurboID. Non-transfected cells were also taken as controls. HeLa cells were treated overnight with oleic acid to enhance lipid droplet surface area.

**Please cite**
Published as Borst Pauwels et al. 2026 \
https://www.biorxiv.org/content/10.64898/2026.01.27.699657v1


```{r load libraries}
library(readxl)
library(ggplot2)
library(dplyr)
library(ggthemes)
library(ggrepel)
library(gridExtra)
```


```{r load data}
bioid1 = read.csv("data/Book1_for_R_with_FFAT_scores.csv")

```

## Wild-type v all other negative-controls unable to bind FFAT motifs
```{r WTvALL, warning=FALSE, fig.width= 10}

Xmin = 2
Xmax = 6 
Ymin = 3
Ymax = 13

AR = (Ymax-Ymin)/(Xmax - Xmin)

bioid1 %>% ggplot(aes(Rest_FC, Rest_logP)) +
   geom_vline(xintercept = 0, alpha = 0.5)+
   geom_rect(xmin = Xmin, xmax = Xmax, ymin = Ymin, ymax = Ymax, alpha = 0.1, fill = 'pink') +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = (4-ffatscore2)/2)) +
   theme_base(12) +
   theme(aspect.ratio = 1,legend.position = 'none') +
   xlab("Fold Change (log2[WT/Rest])") +
   ylab("P-value (-log10)") +
   ggtitle('WT v All negative controls') -> plot1

bioid1 %>% ggplot(aes(Rest_FC, Rest_logP)) +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = 4-ffatscore2)) +
   theme_base(12) +
   theme(aspect.ratio = AR, panel.background = element_rect(fill = '#ffe6e6')) +
   xlim(Xmin, Xmax) +
   ylim(Ymin, Ymax) +
   xlab("") +
   ylab("") +
   geom_text_repel(size = 4, aes(label = Gene.Symbol), max.iter = 100000,max.time = 300) -> plot2

grid.arrange(plot1, plot2, widths = c(3, 3), nrow = 1)
```

Some of the genes don't have any Gene Symbols, I don't want to miss them out just in case.
```{R WithoutName}
#test the bunch with no GeneSymbol
bioid1 %>% 
   filter(is.na(Gene.Symbol)) %>% 
   ggplot(aes(Rest_FC, Rest_logP)) +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = 4-ffatscore2)) +
   theme_base(12) +
   theme(aspect.ratio = 1) +
   xlim(-8, 6) +
   ylim(0, 13) +
   xlab("Fold Change (log2[WT/Rest])") +
   ylab("P-value (-log10)") +
   geom_text_repel(size = 4, aes(label = Accession), max.iter = 100000,max.time = 300)+
   ggtitle('Hits with no Gene Symbol', 'Only accession numbers') 

```

### Print all the hits above the thresholds

```{R ListHits, warning=FALSE}
bioid1 %>% 
   filter(Rest_FC > Xmin & Rest_logP>Ymin) %>% 
   select(Gene.Symbol, Accession,  Rest_FC, Rest_logP, ffatscore) %>% 
   print()

```

## Wild-type v Two other negative controls
```{R WTvSmallerOnes, fig.width= 10}

Xmin = 2
Xmax = 10 
Ymin = 2.5
Ymax = 6

AR = (Ymax-Ymin)/(Xmax - Xmin)

bioid1 %>% ggplot(aes(KD_FC, KD_logP)) +
   geom_vline(xintercept = 0, alpha = 0.5)+
   geom_rect(xmin = Xmin, xmax = Xmax, ymin = Ymin, ymax = Ymax, alpha = 0.05, fill = 'pink') +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = (4-ffatscore2)/2)) +
   theme_base(12) +
   theme(aspect.ratio = 1,legend.position = 'none') +
   xlab("Fold Change (log2[WT/KD mutant])") +
   ylab("P-value (-log10)") +
   geom_text_repel(
      data = bioid1 %>% 
         filter(KD_FC > Xmin & KD_logP>Ymin),
      size = 4, aes(label = Gene.Symbol), max.iter = 100000,max.time = 300) +
   ggtitle('WT v KD mutant',"FFAT scores are inversed")  -> plot21



bioid1 %>% ggplot(aes(TM_FC, TM_logP)) +
   geom_vline(xintercept = 0, alpha = 0.5)+
   geom_rect(xmin = Xmin, xmax = Xmax, ymin = Ymin, ymax = Ymax, alpha = 0.05, fill = 'pink') +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = (4-ffatscore2)/2)) +
   theme_base(12) +
   theme(aspect.ratio = 1,legend.position = 'none') +
   xlab("Fold Change (log2[WT/TM])") +
   ylab("P-value (-log10)") +
   geom_text_repel(
      data = bioid1 %>% 
         filter(KD_FC > Xmin & KD_logP>Ymin),
      size = 4, aes(label = Gene.Symbol), max.iter = 100000,max.time = 300) +
   ggtitle('WT v TM only',"FFAT scores are inversed") -> plot22

grid.arrange(plot21, plot22, nrow =1)




```

## Check FFAT independent interaction partners
WIP



## GO ANALYSIS — WIP
```{r}
#### GO ANALYSIS ####
# BiocManager::install("AnnotationDbi", force = T)
# 
# library(clusterProfiler)
# library(org.Hs.eg.db)  # Replace with your organism (e.g., org.Mm.eg.db for mouse)
# library(AnnotationDbi)
# library(enrichplot)
# 
# # Example: Load your data (ensure it has columns: Gene, FC, logP)
# #dat2 %>% head
# 
# # Convert gene symbols to Entrez IDs
# 
# # Merge with fold change & logP values
# # Order by significance (optional, useful for GSEA)
# 
# # (A) Over-Representation Analysis (ORA)
# # ORA tests if GO terms are enriched in a list of significant genes (e.g., fold change > threshold).
# # Define a cutoff for significance (e.g., FC > 1.5)
# significant_genes <- bioid1$Entrez.Gene.ID[bioid1$Rest_FC > 1.5]
# 
# 
# # Perform GO enrichment for Biological Process (BP)
# go_results_BP <- enrichGO(gene     = significant_genes,
#                        OrgDb    = org.Hs.eg.db,
#                        keyType  = "ENTREZID",
#                        ont      = "BP",   # Options: "BP" (Biological Process), "MF", "CC"
#                        pAdjustMethod = "BH",
#                        pvalueCutoff  = 0.05)
# 
# go_results_MF <- enrichGO(gene     = significant_genes,
#                           OrgDb    = org.Hs.eg.db,
#                           keyType  = "ENTREZID",
#                           ont      = "MF",   # Options: "BP" (Biological Process), "MF", "CC"
#                           pAdjustMethod = "BH",
#                           pvalueCutoff  = 0.05)
# 
# go_results_CC <- enrichGO(gene     = significant_genes,
#                           OrgDb    = org.Hs.eg.db,
#                           keyType  = "ENTREZID",
#                           ont      = "CC",   # Options: "BP" (Biological Process), "MF", "CC"
#                           pAdjustMethod = "BH",
#                           pvalueCutoff  = 0.05)
# 
# 
# plot4 = dotplot(go_results_BP, showCategory=50) + ggtitle("BP")
# plot5 = dotplot(go_results_CC, showCategory=50) + ggtitle("CC")
# plot6 = dotplot(go_results_MF, showCategory=50) + ggtitle("MF")
# 
# grid.arrange(plot4, plot5, plot6, nrow =3 )

```

```{R}
#git config http.postBuffer 524288000
#git pull & git push
#git commit -m "Update files"
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
