---
title: "VAPB interactome"
author: "Birol Cabukusta"
date: "2026-07-28"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F)
```

## 2026

Here we performed TurboID experiments with N-terminal TurboID-tagged VAPB, its point mutant unable to interact with FFAT motifs [K87D, M89D], TM helix only, soluble, free TurboID. Non-transfected cells were also taken as controls. HeLa cells were treated overnight with oleic acid to enhance lipid droplet surface area.

**Please cite:** Published as Borst Pauwels et al. 2026\
<https://www.biorxiv.org/content/10.64898/2026.01.27.699657v1>

```{r load libraries}
library(readxl)
library(ggplot2)
library(dplyr)
library(ggthemes)
library(ggrepel)
library(gridExtra)
library(limma)
library(tibble)
```

```{r load data}
bioid1 = read.csv("data/Book1_for_R_with_FFAT_scores.csv")

```

## Wild-type v controls
Here we compared WT to all negative controls (KD/MD mutant, TM only, TurboID only, and non-transfected cells).

```{r WTvALL, warning=FALSE, fig.width= 10}

Xmin = 2
Xmax = 6 
Ymin = 3
Ymax = 13

AR = (Ymax-Ymin)/(Xmax - Xmin)

bioid1 %>% ggplot(aes(Rest_FC, Rest_logP)) +
   geom_vline(xintercept = 0, alpha = 0.5)+
   geom_rect(xmin = Xmin, xmax = Xmax, ymin = Ymin, ymax = Ymax, alpha = 0.1, fill = 'pink') +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = (4-ffatscore2)/2)) +
   theme_base(12) +
   theme(aspect.ratio = 1,legend.position = 'none') +
   xlab("Fold Change (log2[WT/Rest])") +
   ylab("P-value (-log10)") +
   ggtitle('WT v All negative controls') -> plot1

bioid1 %>% ggplot(aes(Rest_FC, Rest_logP)) +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = 4-ffatscore2)) +
   theme_base(12) +
   theme(aspect.ratio = AR, panel.background = element_rect(fill = '#ffe6e6')) +
   xlim(Xmin, Xmax) +
   ylim(Ymin, Ymax) +
   xlab("") +
   ylab("") +
   geom_text_repel(size = 4, aes(label = Gene.Symbol), max.iter = 100000,max.time = 300) -> plot2

grid.arrange(plot1, plot2, widths = c(3, 3), nrow = 1)
```

Some of the genes don't have any Gene Symbols, I don't want to miss them out just in case.

```{R WithoutName}
#test the bunch with no GeneSymbol
bioid1 %>% 
   filter(is.na(Gene.Symbol)) %>% 
   ggplot(aes(Rest_FC, Rest_logP)) +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = 4-ffatscore2)) +
   theme_base(12) +
   theme(aspect.ratio = 1) +
   xlim(-8, 6) +
   ylim(0, 13) +
   xlab("Fold Change (log2[WT/Rest])") +
   ylab("P-value (-log10)") +
   geom_text_repel(size = 4, aes(label = Accession), max.iter = 100000,max.time = 300)+
   ggtitle('Hits with no Gene Symbol', 'Only accession numbers') 

```

#### Print all the hits above the thresholds

```{R ListHits, warning=FALSE}
bioid1 %>% 
   filter(Rest_FC > Xmin & Rest_logP>Ymin) %>% 
   select(Gene.Symbol, Accession,  Rest_FC, Rest_logP, ffatscore) %>% 
   print()

```

#### Wild-type v two key negative controls

```{R WT_v_individual_controls , fig.width= 10}

Xmin = 2
Xmax = 10 
Ymin = 2.5
Ymax = 6

AR = (Ymax-Ymin)/(Xmax - Xmin)

bioid1 %>% ggplot(aes(KD_FC, KD_logP)) +
   geom_vline(xintercept = 0, alpha = 0.5)+
   geom_rect(xmin = Xmin, xmax = Xmax, ymin = Ymin, ymax = Ymax, alpha = 0.05, fill = 'pink') +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = (4-ffatscore2)/2)) +
   theme_base(12) +
   theme(aspect.ratio = 1,legend.position = 'none') +
   xlab("Fold Change (log2[WT/KD mutant])") +
   ylab("P-value (-log10)") +
   geom_text_repel(
      data = bioid1 %>% 
         filter(KD_FC > Xmin & KD_logP>Ymin),
      size = 4, aes(label = Gene.Symbol), max.iter = 100000,max.time = 300) +
   ggtitle('WT v KD mutant',"FFAT scores are inversed")  -> plot21



bioid1 %>% ggplot(aes(TM_FC, TM_logP)) +
   geom_vline(xintercept = 0, alpha = 0.5)+
   geom_rect(xmin = Xmin, xmax = Xmax, ymin = Ymin, ymax = Ymax, alpha = 0.05, fill = 'pink') +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = (4-ffatscore2)/2)) +
   theme_base(12) +
   theme(aspect.ratio = 1,legend.position = 'none') +
   xlab("Fold Change (log2[WT/TM])") +
   ylab("P-value (-log10)") +
   geom_text_repel(
      data = bioid1 %>% 
         filter(KD_FC > Xmin & KD_logP>Ymin),
      size = 4, aes(label = Gene.Symbol), max.iter = 100000,max.time = 300) +
   ggtitle('WT v TM only',"FFAT scores are inversed") -> plot22

grid.arrange(plot21, plot22, nrow =1)




```

## Check FFAT independent interaction partners
I want to look at non-FFAT interactors: Interactions mediated by MSP domain, but not FFAT motif, medieated by the transmembrane helix only. For this, I used a code generated by ChatGPT. It does linear modeling and adjusted p values.

#### Wild-type v KD
Let's try it first with the wild-type protein


```{R WT_KD_GPT}

sample_columns <- c(
  "WT1", "WT2", "WT3",
  "KD1", "KD2", "KD3"
)

#inspect data
#summary(bioid1[sample_columns])   

#Convert intensity columns to numeric:
bioid1[sample_columns] <- lapply(
  bioid1[sample_columns],
  as.numeric
)

#Filter proteins with insufficient measurements
keep <- rowSums(!is.na(bioid1[c("WT1", "WT2", "WT3")])) == 3 |
        rowSums(!is.na(bioid1[c("KD1", "KD2", "KD3")])) == 3
bioid1_filtered <- bioid1[c(keep),]

#Log₂-transform the intensities
expression_matrix <- as.matrix(
  bioid1_filtered[sample_columns])
expression_matrix <- log2(expression_matrix)

expression_matrix[!is.finite(expression_matrix)] <- NA
wt_columns <- c("WT1", "WT2", "WT3")
kd_columns <- c("KD1", "KD2", "KD3")

keep <- rowSums(!is.na(expression_matrix[, wt_columns, drop = FALSE])) >= 2 &
        rowSums(!is.na(expression_matrix[, kd_columns, drop = FALSE])) >= 2

expression_matrix_clean <- expression_matrix[keep, , drop = FALSE]

# ## Optional: normalise between samples #SKIPPED
# #
# expression_matrix <- normalizeBetweenArrays(
#   expression_matrix,
#   method = "quantile"
# )
# 
# #For proteomics, I often prefer median centring because it is less aggressive:
# sample_medians <- apply(
#   expression_matrix,
#   2,
#   median,
#   na.rm = TRUE
# )
# expression_matrix <- sweep(
#   expression_matrix,
#   2,
#   sample_medians,
#   FUN = "-"
# )

# Create the experimental design
group <- factor(
  c("WT", "WT", "WT", "KD", "KD", "KD"),
  levels = c("WT", "KD")
)

design <- model.matrix(~ 0 + group)
colnames(design) <- levels(group)
#design


#Run the differential-expression analysis
fit <- lmFit(expression_matrix, design)
contrast_matrix <- makeContrasts(
  WT_v_KD = WT - KD,
  levels = design
)

fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2, trend = TRUE, robust = TRUE)

results <- topTable(
  fit2,
  coef = "WT_v_KD",
  number = Inf,
  adjust.method = "BH"
)


results_merged1 = results %>% rownames_to_column('Roster') %>% 
   merge(bioid1_filtered %>% 
            select(Gene.Symbol, ffatscore2) %>% 
            rownames_to_column('Roster'), by = "Roster")

results_merged1 %>% 
   ggplot(aes(logFC, -log10(adj.P.Val))) +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = (4-ffatscore2))) +
   theme_base(12) +
   theme(aspect.ratio = 1,legend.position = 'none') +
   xlab("Fold Change (log2[WT/KD mutant])") +
   ylab("P-value (-log10)") +
   geom_text_repel(
      data = results_merged1 %>% 
         filter(logFC > 1 & log10(adj.P.Val) < -2.5),
      size = 4, aes(label = Gene.Symbol), max.iter = 100000,max.time = 300, nudge_x = 1) +
   ggtitle('WT v KD mutant',"FFAT scores are inversed")  


   

```

#### KD v TM
```{R KD_TM_GPT}
sample_columns <- c(
  "KD1", "KD2", "KD3",
  "TM1", "TM2", "TM3"
)

#Convert intensity columns to numeric:
bioid1[sample_columns] <- lapply(
  bioid1[sample_columns],
  as.numeric
)

#Filter proteins with insufficient measurements
keep <- rowSums(!is.na(bioid1[c("KD1", "KD2", "KD3")])) == 3 |
        rowSums(!is.na(bioid1[c("TM1", "TM2", "TM3")])) == 3

bioid1_filtered <- bioid1[c(keep),]

#Log₂-transform the intensities
expression_matrix <- as.matrix(
  bioid1_filtered[sample_columns])

expression_matrix <- log2(expression_matrix)

expression_matrix[!is.finite(expression_matrix)] <- NA
KD_columns <- c("KD1", "KD2", "KD3")
TM_columns <- c("TM1", "TM2", "TM3")

keep <- rowSums(!is.na(expression_matrix[, KD_columns, drop = FALSE])) >= 2 &
        rowSums(!is.na(expression_matrix[, TM_columns, drop = FALSE])) >= 2

expression_matrix_clean <- expression_matrix[keep, , drop = FALSE]

group <- factor(
  c("KD", "KD", "KD", "TM", "TM", "TM"),
  levels = c("KD", "TM"))

design <- model.matrix(~ 0 + group)
colnames(design) <- levels(group)
#design

#Run the differential-expression analysis
fit <- lmFit(expression_matrix, design)
contrast_matrix <- makeContrasts(
  KD_v_TM = KD - TM,
  levels = design
)

fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2, trend = TRUE, robust = TRUE)

results <- topTable(
  fit2,
  coef = "KD_v_TM",
  number = Inf,
  adjust.method = "BH"
)

results_merged2 = results %>% rownames_to_column('Roster') %>% 
   merge(bioid1_filtered %>% 
            select(Gene.Symbol, ffatscore2) %>% 
            rownames_to_column('Roster'), by = "Roster")


Xmin = -8
Xmax = -3
Ymin = 2.9
Ymax = 4.5

AR = (Ymax-Ymin)/(Xmax - Xmin)

results_merged2 %>% 
   ggplot(aes(logFC, -log10(adj.P.Val))) +
   geom_rect(xmin = Xmin, xmax = Xmax, ymin = Ymin, ymax = Ymax, fill = "pink", alpha = 0.1) +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = (4-ffatscore2))) +
   theme_base(12) +
   theme(aspect.ratio = 0.5,legend.position = 'none') +
   xlab("Fold Change (log2[KD/TM mutant])") +
   ylab("P-value (-log10)") +
   geom_text_repel(
      data = results_merged2 %>% 
         #filter(abs(logFC) > 2 & log10(adj.P.Val) < -2.8),
         filter(logFC > 2 & log10(adj.P.Val) < -2.8),
      size = 4, aes(label = Gene.Symbol), min.segment.length = 0,
      max.iter = 100000,max.time = 300,max.overlaps = Inf) +
   ggtitle('KD v TM mutant',"FFAT scores are inversed")  

results_merged2 %>% 
   ggplot(aes(logFC, -log10(adj.P.Val))) +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = (4-ffatscore2))) +
   theme_base(12) +
   theme(aspect.ratio = AR,legend.position = 'none') +
   xlab("Fold Change (log2[KD/TM mutant])") +
   ylab("P-value (-log10)") +
   xlim(Xmin,Xmax)+
   ylim(Ymin, Ymax) +
   geom_text_repel(size = 4, aes(label = Gene.Symbol), min.segment.length = 0,
      max.iter = 100000,max.time = 300,max.overlaps = Inf) +
   ggtitle('KD v TM mutant',"Focusing on the highlighted section")  


```


#### TM v TurboID-only

List of TM enriched proteins are below.
```{R TM_v_C1}
sample_columns <- c(
  "TM1", "TM2", "TM3",
  "C11", "C12", "C13"
)

#inspect data
#summary(bioid1[sample_columns])   

#Convert intensity columns to numeric:
bioid1[sample_columns] <- lapply(
  bioid1[sample_columns],
  as.numeric
)

#Filter proteins with insufficient measurements
keep <- rowSums(!is.na(bioid1[c("TM1", "TM2", "TM3")])) == 3 |
        rowSums(!is.na(bioid1[c("C11", "C12", "C13")])) == 3

bioid1_filtered <- bioid1[c(keep),]

#Log₂-transform the intensities
expression_matrix <- as.matrix(
  bioid1_filtered[sample_columns]
)
expression_matrix <- log2(expression_matrix)

expression_matrix[!is.finite(expression_matrix)] <- NA
TM_columns <- c("TM1", "TM2", "TM3")
C1_columns <- c("C11", "C12", "C13")

keep <- rowSums(!is.na(expression_matrix[, TM_columns, drop = FALSE])) >= 2 &
        rowSums(!is.na(expression_matrix[, C1_columns, drop = FALSE])) >= 2

expression_matrix_clean <- expression_matrix[keep, , drop = FALSE]

# Create the experimental design
group <- factor(
  c("TM", "TM", "TM", "C1", "C1", "C1"),
  levels = c("TM", "C1")
)

design <- model.matrix(~ 0 + group)
colnames(design) <- levels(group)
#design



#Run the differential-expression analysis
fit <- lmFit(expression_matrix, design)
contrast_matrix <- makeContrasts(
  TM_v_C1 = TM - C1,
  levels = design
)

fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2, trend = TRUE, robust = TRUE)

results <- topTable(
  fit2,
  coef = "TM_v_C1",
  number = Inf,
  adjust.method = "BH"
)


results_merged3 = results %>% rownames_to_column('Roster') %>% 
   merge(bioid1_filtered %>% 
            select(Gene.Symbol, ffatscore2) %>% 
            rownames_to_column('Roster'), by = "Roster")


#plot
##
Xmin = 4
Xmax = 10
Ymin = 2.9
Ymax = 5

AR = (Ymax-Ymin)/(Xmax-Xmin)


results_merged3 %>% 
   ggplot(aes(logFC, -log10(adj.P.Val))) +
   geom_rect(xmin = Xmin, xmax = Xmax, ymin = Ymin, ymax = Ymax, fill = "pink", alpha = 0.1) +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = (4-ffatscore2))) +
   theme_base(12) +
   theme(aspect.ratio = 0.5,legend.position = 'none') +
   xlab("Fold Change (log2[TM/C1 mutant])") +
   ylab("P-value (-log10)") +
   ggtitle('TM v C1',"Focusing on the highlighted section")  



results_merged3 %>% 
   ggplot(aes(logFC, -log10(adj.P.Val))) +
   geom_point(shape = 21, color = 'black', fill = '#1a75ff', aes(size = (4-ffatscore2))) +
   theme_base(12) +
   theme(aspect.ratio = AR,legend.position = 'none') +
   xlab("Fold Change (log2[KD/TM mutant])") +
   ylab("P-value (-log10)") +
   xlim(Xmin,Xmax)+
   ylim(Ymin, Ymax) +
   geom_text_repel(size = 4, aes(label = Gene.Symbol), #max.overlaps = Inf,
      max.iter = 100000,max.time = 300) +
   ggtitle('TM v C1',"Focusing on the highlighted section")  


results_merged3 %>% 
   filter(logFC > 2 & log10(adj.P.Val) < -3) %>% 
   select(Gene.Symbol, logFC, adj.P.Val, ffatscore2)

```

## GO ANALYSIS — WIP

```{r}
#### GO ANALYSIS ####
# BiocManager::install("AnnotationDbi", force = T)
# 
# library(clusterProfiler)
# library(org.Hs.eg.db)  # Replace with your organism (e.g., org.Mm.eg.db for mouse)
# library(AnnotationDbi)
# library(enrichplot)
# 
# # Example: Load your data (ensure it has columns: Gene, FC, logP)
# #dat2 %>% head
# 
# # Convert gene symbols to Entrez IDs
# 
# # Merge with fold change & logP values
# # Order by significance (optional, useful for GSEA)
# 
# # (A) Over-Representation Analysis (ORA)
# # ORA tests if GO terms are enriched in a list of significant genes (e.g., fold change > threshold).
# # Define a cutoff for significance (e.g., FC > 1.5)
# significant_genes <- bioid1$Entrez.Gene.ID[bioid1$Rest_FC > 1.5]
# 
# 
# # Perform GO enrichment for Biological Process (BP)
# go_results_BP <- enrichGO(gene     = significant_genes,
#                        OrgDb    = org.Hs.eg.db,
#                        keyType  = "ENTREZID",
#                        ont      = "BP",   # Options: "BP" (Biological Process), "MF", "CC"
#                        pAdjustMethod = "BH",
#                        pvalueCutoff  = 0.05)
# 
# go_results_MF <- enrichGO(gene     = significant_genes,
#                           OrgDb    = org.Hs.eg.db,
#                           keyType  = "ENTREZID",
#                           ont      = "MF",   # Options: "BP" (Biological Process), "MF", "CC"
#                           pAdjustMethod = "BH",
#                           pvalueCutoff  = 0.05)
# 
# go_results_CC <- enrichGO(gene     = significant_genes,
#                           OrgDb    = org.Hs.eg.db,
#                           keyType  = "ENTREZID",
#                           ont      = "CC",   # Options: "BP" (Biological Process), "MF", "CC"
#                           pAdjustMethod = "BH",
#                           pvalueCutoff  = 0.05)
# 
# 
# plot4 = dotplot(go_results_BP, showCategory=50) + ggtitle("BP")
# plot5 = dotplot(go_results_CC, showCategory=50) + ggtitle("CC")
# plot6 = dotplot(go_results_MF, showCategory=50) + ggtitle("MF")
# 
# grid.arrange(plot4, plot5, plot6, nrow =3 )

```

```{R}
#git config http.postBuffer 524288000
#git pull & git push
#git commit -m "Update files"
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
